---
title: "Assignment 5"
subtitle: "Dimensionless form and types of cycles"
author: "Juan Carlos Villaseñor-Derbez"
date: "`r Sys.Date()`"
header-includes: #allows you to add in your own Latex packages
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
output:
    bookdown::pdf_document2:
      toc: no
      fig_caption: true
      number_sections: no
---

```{r, echo = F}
knitr::opts_chunk$set(echo = F)
suppressPackageStartupMessages({
  library(deSolve)
  library(cowplot)
  library(tidyverse)
})
```


# Convert a model of bacterial growth in a chemostat to dimensionless form

$$
\begin{split}
\frac{dN}{dt} = &\left(\frac{K_{max}C}{K_N +C} \right)N - \frac{FN}{V} \\
\frac{dC}{dt} = &-\alpha\left(\frac{K_{max}C}{K_N +C} \right)N - \frac{FC}{V} + \frac{FC_0}{V}
\end{split}
\label{eq:chemostat}
$$

First we need to define variables that split our variables (bacteria, nutrients, time) into a scalar and its dimensions:

- $N^*$, $C^*$, and $t^*$ are the scalars
- $\hat{N}$, $\hat{C}$, and $\hat{t}$ are the time-independent units such that $N = N^* \times \hat{N}$, $C = C^* \times \hat{C}$, and $t = t^* \times \hat{t}$ represent our measurements.

Therefore, we can re-write Eq \ref{eq:chemostat} as:

$$
\begin{split}
\frac{d(N^*\hat{N})}{d(t^*\hat{t})} = &\left(\frac{K_{max}(C^*\hat{C})}{K_N + (C^*\hat{C})} \right) (N^*\hat{N}) - \frac{F(N^*\hat{N})}{V} \\
\frac{d(C^*\hat{C})}{d(t^*\hat{t})} = &-\alpha\left(\frac{K_{max}(C^*\hat{C})}{K_N + (C^*\hat{C})} \right)(N^*\hat{N}) - \frac{F(C^*\hat{C})}{V} + \frac{FC_0}{V}
\end{split}
$$
We can now multiply both sides by and $\hat{t}$, and divide by $\hat{N}$ and $\hat{C}$ obtaining:

$$
\begin{split}
\frac{dN^*}{dt^*} = & \hat{t}K_{max}\left(\frac{C^*}{\frac{K_N}{\hat{C}} + C^*} \right) N^* - \hat{t}\frac{F}{V} N^* \\
\frac{d(C^*\hat{C})}{d(t^*\hat{t})} = & \frac{-\alpha\hat{t}K_{max}\hat{N}}{\hat{C}}\left(\frac{C^*}{\frac{K_N}{\hat{C}} + C^*} \right)N^* - \hat{t}\frac{F}{V}C^* + \hat{t}\frac{FC_0}{V\hat{C}}
\end{split}
$$

Since $V$ is in units of volume, and $F$ represents a flow (*i.e.* volume over time), we can say that $\hat{t} = \frac{V}{F}$. This is also convenient to later re-write the system. By further setting $\hat{C} = K_N$ and $\hat{N} = \frac{\hat{C}}{\alpha\hat{t}K_{max}} = \frac{\hat{K_N}}{\alpha\hat{t}K_{max}}$ we can re-write the system as:

$$
\begin{split}
\frac{dN}{dt} = &\left(\frac{K_{max}C}{K_N +C} \right)N - \frac{FN}{V} \\
\frac{dC}{dt} = &-\alpha\left(\frac{K_{max}C}{K_N +C} \right)N - \frac{FC}{V} + \frac{FC_0}{V}
\end{split}
$$

# Numerical simulation of a spring with no dampening

The model is described by:

$$
\frac{dx}{dt} = y \\
\frac{dy}{dt} = -Kx
$$

```{r}
spring <- function(t, values, params){
  K <- params[1]
  
  x <- values[1]
  y <- values[2]
  
  dxdt <- y
  dydt <- -K*x
  
  return(list(c(dxdt, dydt)))
}
```

```{r}
K <- 1
params <- c(K)

t <- seq(1, 20, by = 0.1)

Xo <- 2
Yo <- 0.02
values <- c(Xo, Yo)

sol <- deSolve::lsoda(y = values, times = t, parms = params, func = spring) %>% 
  as.data.frame() %>% 
  magrittr::set_colnames(value = c("Time", "X", "Y")) %>% 
  mutate(Xo = Xo,
         Yo = Yo)

Xo <- 2
Yo <- 1
values <- c(Xo, Yo)

sol2 <- deSolve::lsoda(y = values, times = t, parms = params, func = spring) %>% 
  as.data.frame() %>% 
  magrittr::set_colnames(value = c("Time", "X", "Y")) %>% 
  mutate(Xo = Xo,
         Yo = Yo)

Xo <- 1
Yo <- 0.02
values <- c(Xo, Yo)

sol3 <- deSolve::lsoda(y = values, times = t, parms = params, func = spring) %>% 
  as.data.frame() %>% 
  magrittr::set_colnames(value = c("Time", "X", "Y")) %>% 
  mutate(Xo = Xo,
         Yo = Yo)

Xo <- 1
Yo <- 1
values <- c(Xo, Yo)

sol4 <- deSolve::lsoda(y = values, times = t, parms = params, func = spring) %>% 
  as.data.frame() %>% 
  magrittr::set_colnames(value = c("Time", "X", "Y")) %>% 
  mutate(Xo = Xo,
         Yo = Yo)
```


```{r, fig.cap = "Different scenarios for a simple harmonic motion of a spring. A shows the distance (X) through time, and B shows the relationship between distance (X) and speed (Y). Red dots indicate initial conditions, blue dots indicate equilibrium."}
plot1 <- rbind(sol, sol2, sol3, sol4) %>% 
  mutate(Case = paste0("Xo = ", Xo, "; Yo = ", Yo)) %>% 
  ggplot(aes(x = Time, y = X, color = Case)) +
  geom_line() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none")

plot2 <- rbind(sol, sol2, sol3, sol4) %>% 
  mutate(Case = paste0("Xo = ", Xo, "; Yo = ", Yo)) %>% 
  ggplot(aes(x = X, y = Y, color = Case)) +
  geom_point() +
  scale_color_brewer(palette = "Set1") +
  theme(strip.background = element_blank()) +
  coord_fixed()

bottom <- cowplot::plot_grid(plot2, NULL, ncol = 2)

cowplot::plot_grid(plot1, plot2, labels = "AUTO", ncol = 1)
```

# Numerical solutions for the non-linear Lotka-Volterra predator-prey model

\begin{equation}
\begin{split}
\frac{dN}{dt} &= r N - a N P \\	 
\frac{dP}{dt} &= c a N P - m P
\end{split}
\end{equation}

```{r}
lotka <- function(t, values, params){
  # Extract parameters
  r <- params[1]
  a <- params[2]
  c <- params[3]
  m <- params[4]
  f <- params[5]

  # Extract state variables
  N <- values[1]
  P <- values[2]
  
  dNdt <- (r*N) - (a*N*P)
  dPdt <- (a*c*N*P) - (m*P)
  
  return(list(c(dNdt, dPdt)))
}
```

```{r}
get_plot <- function(r, a, c, m, N0, P0){
  
  params <- c(r, a, c, m)
  values <- c(N0, P0)
  
  # Time
  t <- seq(0, 50, by = 0.05)
  
  plot <- deSolve::lsoda(y = values, func = lotka, times = t, parms = params) %>% 
    as.data.frame() %>% 
    magrittr::set_colnames(value = c("Time", "N", "P")) %>% 
    ggplot(aes(x = N, y = P)) +
    geom_point() +
    ggtitle(paste0("r = ",r, ", a = ", a, "\nc = ", c, ", m = ", m)) +
    cowplot::theme_cowplot() +
    geom_point(aes(x = N0, y = P0), color = "red", size = 4) +
    geom_point(aes(x = m/(c*a), y = r/a), color = "blue", size = 4) +
    geom_hline(yintercept = r/a, linetype = "dashed") +
    geom_vline(xintercept = m/(c*a), linetype = "dashed") +
    scale_x_continuous(limits = c(0, 12)) +
    scale_y_continuous(limits = c(0, 4))
  
  return(plot)
}
```

```{r, fig.height = 9, fig.width = 6, fig.cap = "Combination of parameters and initial conditions for a Lotka-Volterra system with no density dependance and a Type-I functional response."}
case1 <- get_plot(r = 1, a = 1, c = 0.1, m = 0.1, N0 = 2, P0 = 1)
case2 <- get_plot(r = 1, a = 1, c = 0.1, m = 0.1, N0 = 4.9, P0 = 1)
case3 <- get_plot(r = 1, a = 1, c = 0.1, m = 0.1, N0 = 2, P0 = 2)
case4 <- get_plot(r = 0.5, a = 1, c = 0.1, m = 0.1, N0 = 2, P0 = 1)
case5 <- get_plot(r = 1, a = 0.5, c = 0.1, m = 0.1, N0 = 4.9, P0 = 1)
case6 <- get_plot(r = 1, a = 1, c = 0.05, m = 0.1, N0 = 2, P0 = 2)

cowplot::plot_grid(case1, case2, case3, case4, case5, case6, ncol = 2)
```





































