---
title: "Assignment 4"
subtitle: "Separation of Time Scales: Rapid Equilibrium Approximation"
author: "Juan Carlos Villaseñor-Derbez"
date: "February 8, 2018"
header-includes: #allows you to add in your own Latex packages
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
output:
    bookdown::pdf_document2:
      toc: no
      fig_caption: true
      number_sections: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 3,
                      fig.width = 5,
                      fog.pos = "h",
                      fig.split = "center")

suppressPackageStartupMessages({
  library(deSolve)
  library(rootSolve)
  library(tidyverse)
})
```

# Defining the system

The system has the following form:

$$
A \hspace{3mm} \overset{\xrightarrow{k_1}}{\underset{\xleftarrow{k_2}}{}} \hspace{3mm} B \hspace{3mm} \overset{\xrightarrow{k_3}}{\underset{\xleftarrow{k_4}}{}} \hspace{3mm}C
$$

And the system of ODE's that define it, is given by:

\begin{equation}
\begin{split}
  \frac{da}{dt} &= (-k_1 * a) + (k_2 * b) \\
  \frac{da}{dt} &= (k_1 * a) + (k_4 * c) - (k_2 * b) - (k_3 * b) \\
  \frac{da}{dt} &= (k_3 * b) - (k_4 * c)
\end{split} \label{eq:system}
\end{equation}

With rate constants $k_1=0.05$, $k_2=0.005$, $k_3=0.7$, and $k_4=0.4$.

# Numerical solution

```{r}
# Define a function
system <- function(t, values, params){
  # Extract parameters
  k1 <- params[1]
  k2 <- params[2]
  k3 <- params[3]
  k4 <- params[4]
  
  # Extract state variables
  a <- values[1]
  b <- values[2]
  c <- values[3]
  
  # Calculate variables
  dadt <- (-k1 * a) + (k2 * b)
  dbdt <- (k1 * a) + (k4 * c) - (k2 * b) - (k3 * b)
  dcdt <- (k3 * b) - (k4 * c)
  
  # Return results
  return(list(c(dadt, dbdt, dcdt)))
}
```

```{r}
# Define timesteps
t <- seq(0, 50)

# Define initial conditions
a <- 1.5
b <- 3
c <- 2
values <- c(a, b, c)

# Define parameters
k1 <- 0.05
k2 <- 0.005
k3 <- 0.7
k4 <- 0.4
params <- c(k1, k2, k3, k4)

# Numerical solution
full <- lsoda(y = values, times = t, func = system, parms = params) %>% 
  as.data.frame() %>% 
  magrittr::set_colnames(value = c("time", "A", "B", "C")) %>% 
  gather(Letter, Value, -time) %>% 
  mutate(Model = "Full")
```

# Visualizing the transient and steady-state behavior of the system

```{r, fig.cap = "Transient and steady-state behavior of state variables A, B, and C"}
ggplot(full, aes(x = time, y = Value, color = Letter)) +
  geom_line(size = 1) +
  cowplot::theme_cowplot() +
  scale_color_brewer(palette = "Set1")
```


# Separation of time-scales by the *rapid equilibrium assumption*

From the Eq \ref{eq:system} we can see that the main processes are between at A-B and B-C. However, they occur at different timescales, governed by the $k_i$ coefficients.

For A-B, the timescale is given by:

$$
\frac{1}{k_1 + k_2} \\
= \frac{1}{0.05 + 0.005} \\
= \frac{1}{0.055}
$$

And the changes between B-C are given by:

$$
\frac{1}{k_3 + k_4} \\
= \frac{1}{0.7 + 0.4} \\
= \frac{1}{1.1}
$$

Since the conversions $B \hspace{3mm} \overset{\xrightarrow{}}{\underset{\xleftarrow{}}{}} \hspace{3mm}C$ have a smaller time constant than the conversion $A \hspace{3mm} \overset{\xrightarrow{}}{\underset{\xleftarrow{}}{}} \hspace{3mm} B \hspace{3mm}$ (*i.e.* $\frac{1}{k_1 + k_2}<<\frac{1}{k_3 + k_4}$), indicating that $B$ and $C$ reach an equilibrium faster than $A$. We can pool $B$ and $C$ together, and focus on how their pooled concentrations and $A$ interact. From Eq \ref{eq:system}, we have that

$$\frac{\tilde{c}(t)}{\tilde{b}(t)} = \frac{k_3}{k_4}$$

which in turn is

\begin{equation}
\label{eq:pool_cb}
\tilde{c}(t)= \tilde{b}(t) \frac{k_3}{k_4}
\end{equation}

This allows us to reduce the system to a pooled B and C group that converts to and from A. Let $\tilde{d}(t)$ be the total concentration in the pool $B+C$ (*i.e.* $\tilde{d}(t) = \tilde{b}(t) + \tilde{c}(t)$). The fractions of $B$ and $C$ are fixed by the equilibrium ratio. Thus, combining this with Eq \ref{eq:pool_cb}, we obtain:

$$
\begin{split}
\tilde{d}(t) &= \tilde{b}(t) + \tilde{c}(t)\\
 &= \tilde{b}(t) + \tilde{b}(t) \frac{k_3}{k_4} \\
 &= \tilde{b}(t)\frac{k_4+k_3}{k_4}
\end{split}
$$

And solving for $\tilde{b}(t)$ gives:

\begin{equation}
\label{eq:tilde_b}
\tilde{b}(t) = \tilde{d}(t)\frac{k_4}{k_4+k_3}
\end{equation}

And

\begin{equation}
\label{eq:tilde_c}
\tilde{c}(t) = \tilde{d}(t) - \tilde{b}(t) = \tilde{d}(t)\frac{k_3}{k_4+k_3}
\end{equation}

\clearpage

The pool $D$ converts to and from $A$ at rate $k_1\tilde{a}(t)-k_2\tilde{b}(t)$:

\begin{equation}
\begin{split}
\frac{d}{dt}\tilde{d}(t) &= k_1\tilde{a}(t)-k_2\tilde{b}(t) \\
&= k_1\tilde{a}(t)-k_2\left(\frac{k_4}{k_4+k_3}\tilde{d}(t)\right) \\
&= k_1\tilde{a}(t)-\frac{k_2k_4}{k_4+k_3}\tilde{d}(t)
\label{eq:tilde_d}
\end{split}
\end{equation}

We can use Eq \ref{eq:tilde_d} along with $\frac{d}{dt}\tilde{a}(t) = (-k_1+k_2)\tilde{d}(t)$ to calculate the long-term equilibrium of $A$ and $D$, and then use the algebraic expressions \ref{eq:tilde_b} and \ref{eq:tilde_c} to calculate $B$ and $C$ at each time step.

# Numerical solution of reduced model

```{r}
# Define a function for the reduced system
reduced_system <- function(t, values, params){
  # Extract parameters
  k1 <- params[1]
  k2 <- params[2]
  k3 <- params[3]
  k4 <- params[4]
  
  # Extract state variables
  a <- values[1]
  d <- values[2]
  
  # Calculate rates
  dadt <- (-k1 * a) + (k2 * d)
  dddt <- (k1 * a) - (k2*((k4)/(k4+k3))*d)
  
  return(list(c(dadt, dddt)))
}
```

```{r}
# Define new initial conditions
d <- b + c
values <- c(a, d)

# Numerical solution
reduced <- lsoda(y = values, times = t, func = reduced_system, parms = params) %>% 
  as.data.frame() %>% 
  magrittr::set_colnames(value = c("time", "A", "D")) %>% 
  mutate(B = D*(k4/(k4+k3)),
         C = D*(k3/(k4+k3))) %>% 
  gather(Letter, Value, -time) %>% 
  mutate(Model = "Reduced")
```

\clearpage

```{r, fig.cap = "Comparision of transient and steady-state behavior of state variables A, B, and C between the full model and the reduced model."}
rbind(full, reduced) %>% 
  ggplot(aes(x = time, y = Value, color = Letter, linetype = Model)) +
  geom_line(size = 1) +
  cowplot::theme_cowplot() +
  scale_color_brewer(palette = "Set1")
```
























