---
title: "Lab 9 notes"
author: "Juan Carlos Villaseñor-Derbez"
date: "9 de marzo de 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Types of models

Time | Discrete stage | Continuous stage |
-----|----------|------------------|
Discrete time | Matrix Model | Integral Projection Model
Continuous time | ODEs, and Delayed Diffeential Equations | PDEs



EEMB 595 TE

PSPM Analysis




Three ways to think about space:

Time | Spatially Implicit | Spatially Explicit | Spatially realistic
-----|--------------------|--------------------|---------------------
Discrete Time | Patch Model | Lattice Models / Celular automata | Properties of space matter
Continuous Time | Patch Models / Levin | PDEs / reaction-diffusion type model | 


**Levin metapopulation model**

$p$ = fraction of patches occupied by a species

$\frac{dp}{dt} = cp(1-p)-mp$

At aquilibrium ($\frac{dp}{dt} = 0$): $p^* = 1-\frac{m}{c}$


**Reaction Diffusion**

$\frac{\partial N(x, t)}{\partial t} = D \frac{\partial^2}{\partial x^2} + rN(x, t)(1-\frac{N(x,t)}{k}$

This would be great for a 1-D marine reserve model

## Advection diffusion model from the class

```{r}
require(ReacTran)

# Establish the grid, using the setup.grid.1D()  function, and supply values for the parameters.
N = 100         # Number of grid cells
xgrid = setup.grid.1D(x.up = 0, x.down = 1, N = N)      # Between 0 and 1
x = xgrid$x.mid     # Midpoints of grid cells
D = 1e-4        # Diffusion coefficient
v = 0         # Advection velocity
r = 0.5         # per-capita growth rate  (make non-zero to add reaction component)
k = 150

# Construct the function that defines the diffusion-advection equation.
Reaction_Diffusion = function(t, Y, parms) {
  tran=tran.1D(C=Y,C.up=0,C.down=0,D=D,dx= xgrid)$dC
  growth=r*Y*(1-(Y/k))
  list(dY = tran+growth)
}

# Initialize the concentration on the grid.
Yini = rep(100,N) # Initial concentration = 0
# Yini[50] = 100 # Except in the second cell

# Now run the calculation for five time units, with a time step of 0.01.
times = seq(from = 0, to = 50, by = 0.01)
out = ode.1D(y = Yini, times = times, func = Reaction_Diffusion, parms = NULL, dimens = N)

# Plot the initial concentration spike and the subsequent concentration distributions 
# at intervals of 50 time steps.
plot(x,out[1,2:(N+1)], type = "l", lwd = 1, xlab = "x", ylab = "Y")
for(i in seq(2, length(times), by = 50)) lines(x, out[i, 2:(N+1)], col = colors()[i])
```




